---
title: "Creating a Movie Recommendation System"
subtitle: "Report for HarvardX PH125.9x: Data Science: Capstone"
author: "Samuel Bates"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
require(gridExtra)

load('ml-10M100K/movielens_report.RData')
```

# Overview

In this report, I describe an exploration of a dataset of movie ratings made by many
people over the course of several years, and the construction of a model for
predicting ratings for specific movies by specific people. To test the model,
I have kept 10% of the ratings out of the model construction. These ratings
will be used to determine how well the model predicts ratings.

# Exploration

The dataset contains 10,000,054 ratings of movies by people (hereinafter referred
to as *users*). There are 10,677
different movies rated and 69,878 different users. The ratings were taken
between January 9, 1995 and January 5, 2009.
The dataset is provided as two files, each containing a table:

- `ratings.dat`, containing 10,000,054 observations, each with 4 variables:
`userId`, `movieId`, `rating` (a number between 0 and 5), and `timestamp`;
- `movies.dat`, containing 10,681 observations, each with 3 variables: `movieId`,
`title` (which contains the title and year released), and `genres` (which
contains a list of genres separated by `|` symbols).

Tables 1 and 2 show the results of counting how many movies each user rated.
Half of all users rated 69 or fewer movies, and 9 out of 10 rated 335 or
fewer movies. Only about 1% of the users rated more than 1000 movies.

```{r user-stats, echo=FALSE}
percents <- ratings %>% 
  group_by(userId) %>% 
  summarize(num_ratings = n()) %>% 
  pull(num_ratings) %>%
  quantile(probs = seq(0, 1, .1))
knitr::kable(t(percents), 
             caption = "Number of movies rated by users")

high_percents <- ratings %>% 
  group_by(userId) %>% 
  summarize(num_ratings = n()) %>% 
  pull(num_ratings) %>%
  quantile(probs = seq(.9, 1, .01))
knitr::kable(t(high_percents), 
             caption = "Number of movies rated by high-volume users")
rm(percents, high_percents)
```


# Methods and Analysis

A model for predicting ratings must be naturally based on which user 
is rating the movie
and which movie is being rated. Figures 1 and 2 show this.

```{r user-movie-correlation, echo=FALSE, fig.height=3}
movie_stats <- ratings %>% group_by(movieId)

mean_movie_ratings <- movie_stats %>%
  summarize(num_ratings = n(), mean_rating = mean(rating))

# TODO: Add axis labels and title. Include corrrelation value in title.
movie_effect <- mean_movie_ratings %>%
  ggplot(aes(num_ratings, mean_rating)) + geom_point() + scale_x_log10()

user_stats <- ratings %>% group_by(userId)

mean_user_ratings <- user_stats %>%
  summarize(num_ratings = n(), mean_rating = mean(rating))

# TODO: Add axis labels and title. Include corrrelation value in title.
user_effect <- mean_user_ratings %>%
  ggplot(aes(num_ratings, mean_rating)) + geom_point() + scale_x_log10()

grid.arrange(movie_effect, user_effect, ncol = 2)
rm(mean_movie_ratings, movie_effect, mean_user_ratings, user_effect)
```
# Results



# Conclusion

